Formalizing Stack Safety as a Security Property

### 2023-02-02 submission5/core.tex
work on 3 and a few small things in 4

In this section we consider a simple setting with calls, returns, and private allocations,
on a RISC-V machine with a PIPE tagging system. In this system a value is a pair of a
word (the ``payload'') and a tag drawn from some set \(\mathcal{T}\).
%\(\mathcal{V} ::= \mathcal{W} \times \mathcal{T}\),
%for some type  of tags to be elaborated later, with \(|(w,t)| = w\).
Figure \ref{fig:RISCVregs} gives an example division of the RISC-V registers into
security classes.
Figure \ref{fig:main} gives the C code and corresponding compiled RISC-V code
for a function {\tt main}. Code is compiled using the standard RISC-V
calling conventions~\cite{??}; the stack grows towards lower addresses. We assume that
{\tt main} begins at address 0 and its callee {\tt f} at address 100.

{\tt main} takes a secret as its argument and initializes a variable to contain
potentially sensitive data. We define an observable event to be a store
to a special address, called {\tt out}.
{\tt main} calls another function {\tt f},
and afterward may decide to output its secret based on the contents of
its sensitive data. In this case, since {\tt sensitive} is initialized to 0,
it should not---instead it will output the results of {\tt f}.

@@

      \}}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \begin{tabular}{r l | l}
      \labeledrow{0:}{addi sp,sp,-20}{\(\mathbf{alloc} ~ (-20,20)\)}
      \labeledrow{4:}{sd ra,12(sp)}{}
      \labeledrow{8:}{sw a1,8(sp)}{}
      \labeledrow{12:}{sw zero,4(sp)}{}
      \labeledrow{16:}{jal f,ra}{\(\mathbf{call} ~ \mathtt{f} ~ \emplist ~ \emplist\)}
      \labeledrow{20:}{sw a0,0(sp)}{}
      \labeledrow{24:}{lw a4,4(sp)}{}
      \labeledrow{28:}{li a5,42}{}
      \labeledrow{32:}{bne a4,a5,L1}{}
      \labeledrow{36:}{lw a0,8(sp)}{}
      \labeledrow{40:}{sw a0,out}{}
      \labeledrow{44:}{j L2:}{}
      \labeledrow{L1, 48:}{lw a0,0(sp)}{}
      \labeledrow{52:}{sw a0,out}{}
      \labeledrow{L2, 56:}{ld ra,12(sp)}{}
      \labeledrow{60:}{addi sp,sp,20}{\(\mathbf{dealloc} ~ (0,20)\)}
      \labeledrow{64:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{stacklayout.png}
  \end{subfigure}

\caption{Example: Main}
\label{fig:main}
\end{figure}

Suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.
%
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.)
%
In Figure \ref{subfig:WBCF2}, a different attack violates WBCF, this time
by returning to the correct program counter but with the wrong stack pointer.
\footnote{We pad the last two variants with {\tt nop}s just so that all the
snippets have the same length, which simplifies some later explanations.}

\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{sw a4,out}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} directly}
    \label{subfig:direct}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{mov a4,a0}{}
      \labeledrow{108:}{sw zero,-4(sp)}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} indirectly}
    \label{subfig:indirect}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{li a5,42}{}
      \labeledrow{104:}{sw a5,4(sp)}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking {\tt sensitive}}
    \label{subfig:integrity}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi ra,ra,16}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking control flow}
    \label{subfig:WBCF}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi sp,sp,8}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Relative address attack}
    \label{subfig:WBCF2}
  \end{subfigure}

  \caption{Assembly code alternatives for {\tt f} as an attacker.
  }
  \label{fig:f}
\end{figure}

@@

Recall\apt{can't; we've lost these} that our security semantics attach a security context to the machine state,
which consists of a view \(V\) and a stack \(\sigma\) of pending activations' views.
(See \cref{sec:machine} for details.)
Figure \ref{fig:exec1} shows how the security context evolves over the first few
steps of the program.  (The rules governing this evolution are formalized in \cref{fig:advops}.)
It begins at the start of {\tt main}, where the program counter (\(\PCname\)) is zero,
and with the stack pointer (\(\SP\)) at address 1000.
State transitions are numbered and labeled with a list of security operations, written
\(\downarrow \overline{\psi}\) between steps.

Step 1 allocates a word each for {\tt secret}, {\tt sensitive}, and {\tt res}, as well
as two words for the return address. This will have the
effect of marking those bytes \(\object\), assuming they were previously
\(\unsealed\). (We use \(V\llbracket\cdot\rrbracket\) to denote updates to \(V\).)
%

@@

\begin{figure*}
  \begin{tabular}{|r|r||l|r}
    \cline{1-3}
    \(\PCname\) & \(\SP\) & Context &
    \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox
        \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\pubbox} \!\! \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a1}}{\pubbox} ~ \stackrel{\mathtt{a2}}{\freebox}
      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)} \\
    \cline{1-3}
    0 & 1000 & \(V_0, \emplist\)
    \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{\(1 \Big\downarrow [\mathbf{alloc} ~ (-20,20)]\)}} & \\
    \multicolumn{3}{l}{} &
    \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\objbox} \!\! \objbox \objbox \objbox \objbox}_\object
      \! \underbrace{\pubbox \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a1}}{\pubbox} ~ \stackrel{\mathtt{a2}}{\freebox}
      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)}
    \\
    \cline{1-3}
    4 & 980 & \(V_1 = V_0 \llbracket 980..999 \mapsto \object\rrbracket, \emplist\) &
    \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{2-4 \(\Big\downarrow \emplist\)}} \\ \multicolumn{3}{l}{} \\
    \cline{1-3}
    16 & 980 & \(V_1, \emplist\) & \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{\(5 \Big\downarrow [\mathbf{call} ~ 100 ~ \emplist]\)}} & \\
    \multicolumn{3}{l}{} &
    \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\sealbox} \!\! \sealbox \sealbox \sealbox \sealbox}_\sealed
      \! \underbrace{\pubbox \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a1}}{\freebox} ~ \stackrel{\mathtt{a2}}{\freebox}
      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)}
    \\
    \cline{1-3}
    100 & 980 & \(V_2 = V_1 \llbracket 980..999 \mapsto \sealed, \mathtt{a0} \mapsto \unsealed\rrbracket,[V_1]\) & \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{6-8 \(\Big\downarrow \emplist\)}} \\ \multicolumn{3}{l}{} \\
    \cline{1-3}
    112 & 980 & \(V_2,[V_1]\) \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{\(9 \Big\downarrow [\mathbf{return}]\)}} & \\
    \multicolumn{3}{l}{} & \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\objbox} \!\! \objbox \objbox \objbox \objbox}_\object
      \! \underbrace{\pubbox \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a1}}{\pubbox} ~ \stackrel{\mathtt{a2}}{\freebox}
      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)}
    \\
    \cline{1-3}
    20 & 980  & \(V_1, \emplist\) &
    \\
    \cline{1-3}
    \multicolumn{2}{l}{} \\
  \end{tabular}
\caption{Execution up through the return from {\tt f}}
\label{fig:exec1}
\end{figure*}
%
At step 5, the current principal's record is pushed onto the inactive list.
Its return target is the return address of the call,
and the stack pointer target is the stack pointer at the moment of call.
The callee's view is updated from the caller's such that all \(\object\) locations
become \(\sealed\). (In more complex settings, objects that are intentionally passed
to the callee will not get sealed, but for now we assume no sharing of memory between activations.)
For registers, \(\overline{\reg_{args}}\) tells us which registers are used as arguments, 
in this case just {\tt a0}. \apt{We've lost the explanation of the arguments to call, and they are used inconsistent now too, so this is utterly mysterious.}
These are mapped to \(\public\), while any non-argument, caller-saved
registers remain mapped to \(\unsealed\). All callee-save registers remain \(\sealed\) for all calls.
At step 9, {\tt f} returns, and the topmost inactive view, that of {\tt main}, is restored.

We now show how this security semantics can be used define notions of confidentiality,
integrity, and correct control flow in such a way that many classes of
bad behavior, including the attacks in Figure~\ref{fig:f}, are
are detected as security violations.

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.

Consider the attack in Figure \ref{subfig:WBCF}: the attacker adds
16 to the return address and then returns, thus bypassing the {\tt if}-test in the code and outputting
{\tt secret}. Before the call, the program counter is 16 and the stack pointer is 980,
so we define a predicate on states that should hold just after the return:
\(\ret\ \mach \triangleq \mach[\PCname] = 20 \wedge \mach[\SP] = 980\).
%
We can identify the point just after the return (if a return occurs)
as the first state in which the pending call stack is smaller than it was
just after the call.
WBCF requires that if \(\mach\) is the state at that point, then \(\ret ~ \mach\) holds.
%For nested calls, where the pending stack is initially larger, the same principle
%applies: \(\ret ~ \mach\) must hold the next time the pending stack is the same size or smaller.

To see why \(\wbcf\) also cares about the stack pointer, consider the attack in
Figure \ref{subfig:WBCF2}. Here the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}
\end{center}

@@

\noindent
{\tt main}'s attempt to read {\tt sensitive} will instead
read part of the return address, and its attempt to output
{\tt res} will instead output {\tt secret}!
% Even absent other
% kinds of data protection, the stack pointer {\it must} be restored
% for the program to behave predictably.

@@

Like WBCF, stack integrity defines a condition at the call that must hold upon
return. This time the condition applies to all of the memory that a function has
allocated. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labels it \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, it can rely on any \(\sealed\) elements
to be identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

@@

To characterize this situation in a property, we borrow the idea of \emph{variant} states
from the theory of non-interference. For a set of elements \(\components\),
a pair of states \(\mach\) and \(\nach\) are {\em \(\components\)-variants} if
their values an only disagree on elements in \(\components\). We further require that
the respective values in \(\components\) be \emph{compatible}, meaning that even though
their payloads may differ, their tags (or whatever other data is important to
the machine's security enforcement mechanism) must be identical.
We say that the elements of \(\components\) are \emph{irrelevant}
in \(\mach\) if they can be replaced by compatible values without changing the
observable behavior of the machine. All other elements are \emph{relevant}.
Variants and relevance are defined formally in \cref{sec:props}.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{variants.png}
  \caption{Integrity Violation\apt{Needs a machine-drawn version, as do the rest.}}
  \label{fig:variant}
\end{figure}

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to WBCF, we will define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We once again require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43 (and keeping its tag, \(t_2'\), unchanged.) As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

@@

@@

The system we've modeled so far has been very simple, but our properties
support an important dimension of generality: adding features. We will now
briefly introduce the features of our primary testing machine.

>>>>

In this section we introduce our security property definitions by means
of small code examples, using a simplified set of security-relevent operations for
or calls, returns, and private allocations.
% In this system a value is a pair of a
%word (the ``payload'') and a tag drawn from some set \(\mathcal{T}\).
%\(\mathcal{V} ::= \mathcal{W} \times \mathcal{T}\),
%for some type  of tags to be elaborated later, with \(|(w,t)| = w\).
Figure \ref{fig:main} gives C code and corresponding compiled 32-bit RISC-V code
for a function {\tt main}, which  
takes an argument {\tt secret} and initializes a local variable {\tt sensitive} to contain
potentially sensitive data. 
Then {\tt main} calls another function {\tt f},
and afterward performs a test on {\tt sensitive} to decide whether
to output {\tt secret}.  Since {\tt sensitive} is initialized to 0,
the test should fail, and {\tt main} should instead output the return value of {\tt f}.
Output is performed by writing to the special global {\tt out},
and we assume that such writes are the only observable events in the system.

The C code is compiled using the standard RISC-V calling conventions~\cite{RISC-V-CC}.
In particular, the first function argument and the function
return value are both passed in {\tt a0}.  
Memory is byte-addressed and the stack grows towards
lower addresses. We assume that {\tt main} begins at address 0 and its callee {\tt f} at address 100.

@@

      \}}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \begin{tabular}{r l | l}
      \labeledrow{0:}{addi sp,sp,-20}{\(\mathbf{alloc} ~ (-20,20)\)}
      \labeledrow{4:}{sd ra,12(sp)}{}
      \labeledrow{8:}{sw a0,8(sp)}{}
      \labeledrow{12:}{sw zero,4(sp)}{}
      \labeledrow{16:}{jal f,ra}{\(\mathbf{call} ~ \emplist \)}
      \labeledrow{20:}{sw a0,0(sp)}{}
      \labeledrow{24:}{lw a4,4(sp)}{}
      \labeledrow{28:}{li a5,42}{}
      \labeledrow{32:}{bne a4,a5,L1}{}
      \labeledrow{36:}{lw a0,8(sp)}{}
      \labeledrow{40:}{sw a0,out}{}
      \labeledrow{44:}{j L2:}{}
      \labeledrow{L1, 48:}{lw a0,0(sp)}{}
      \labeledrow{52:}{sw a0,out}{}
      \labeledrow{L2, 56:}{ld ra,12(sp)}{}
      \labeledrow{60:}{addi sp,sp,20}{\(\mathbf{dealloc} ~ (0,20)\)}
      \labeledrow{64:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{stacklayout.png}
  \end{subfigure}

\caption{Example: {\tt main}\apt{Needs machine-set diagram.}}
\label{fig:main}
\end{figure}

%Figure \ref{fig:RISCVregs} gives an example division of the RISC-V registers into
%security classes.  APT: Don't think we need this figure.

We now consider how {\tt f} might misbehave and violate desirable
stack safety properties associated with {\tt main}. To put the violations in a
security framework, suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.
%
Leakage is most obviously viewed as a violation of {\tt main}'s {\it confidentiality}.
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.)
%
In Figure \ref{subfig:WBCF2}, a different attack violates WBCF, this time
by returning to the correct program counter but with the wrong stack pointer.
\footnote{We pad the last two variants with {\tt nop}s just so that all the
snippets have the same length, which simplifies some later explanations.}

\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{sw a4,out}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} directly}
    \label{subfig:direct}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{mov a4,a0}{}
      \labeledrow{108:}{sw zero,-4(sp)}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} indirectly}
    \label{subfig:indirect}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{li a5,42}{}
      \labeledrow{104:}{sw a5,4(sp)}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking {\tt sensitive}}
    \label{subfig:integrity}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi ra,ra,16}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking control flow}
    \label{subfig:WBCF}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi sp,sp,8}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking stack pointer integrity}
    \label{subfig:WBCF2}
  \end{subfigure}

  \caption{Example: assembly code alternatives for {\tt f} as an attacker.
  }
  \label{fig:f}
\end{figure}

@@

Recall\apt{make sure this was said in previous section!} that our security semantics attach a security context to the machine state,
which consists of a view \(V\) and a stack \(\sigma\) of pending activations' views.
Figure \ref{fig:exec1} shows how the security context evolves over the first few
steps of the program.  (The formal details of the security semantics are described in
\cref{sec:machine}, and the context evolution rules are formalized in \cref{fig:basicops}.)
Execution begins at the start of {\tt main}, where the program counter (\(\PCname\)) is zero,
and with the stack pointer (\(\SP\)) at address 1000.
State transitions are numbered and labeled with a list of security operations, written
\(\downarrow \overline{\psi}\) between steps.

The initial view \(V_0\) maps all stack addresses below \(\SP\) to \(\unsealed\) and the remainder of
memory to \(\public\). The sole used argument register, {\tt a0}, is mapped to \(public\) \apt{why not \(unsealed\)?};
other caller-save registers are mapped to \(\unsealed\) and callee-save registers to \(\sealed\). 
Step 1 allocates a word each for {\tt secret}, {\tt sensitive}, and {\tt res}, as well
as two words for the return address. This has the
effect of marking those bytes \(\object\). 
(We use \(V\llbracket\cdot\rrbracket\) to denote updates to \(V\).)
%

@@

\begin{figure*}
  \begin{tabular}{|r|r||l|r}
    \cline{1-3}
    \(\PCname\) & \(\SP\) & Context &
    \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox
        \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\pubbox} \!\! \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a0}}{\pubbox} ~ \stackrel{\mathtt{a1}}{\freebox}
%      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)} \\
    \cline{1-3}
    0 & 1000 & \(V_0, \emplist\)
    \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{\(1 \Big\downarrow [\mathbf{alloc} ~ (-20,20)]\)}} & \\
    \multicolumn{3}{l}{} &
    \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\objbox} \!\! \objbox \objbox \objbox \objbox}_\object
      \! \underbrace{\pubbox \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a0}}{\pubbox} ~ \stackrel{\mathtt{a1}}{\freebox}
%      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)}
    \\
    \cline{1-3}
    4 & 980 & \(V_1 = V_0 \llbracket 980..999 \mapsto \object\rrbracket, \emplist\) &
    \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{2-4 \(\Big\downarrow \emplist\)}} \\ \multicolumn{3}{l}{} \\
    \cline{1-3}
    16 & 980 & \(V_1, \emplist\) & \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{\(5 \Big\downarrow [\mathbf{call} ~ 100 ~ \emplist]\)}} & \\
    \multicolumn{3}{l}{} &
    \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\sealbox} \!\! \sealbox \sealbox \sealbox \sealbox}_\sealed
      \! \underbrace{\pubbox \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a0}}{\freebox} ~ \stackrel{\mathtt{a1}}{\freebox}
%      ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)}
    \\
    \cline{1-3}
    100 & 980 & \(V_2 = V_1 \llbracket 980..999 \mapsto \sealed, \mathtt{a0} \mapsto \unsealed\rrbracket,[V_1]\) & \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{6-8 \(\Big\downarrow \emplist\)}} \\ \multicolumn{3}{l}{} \\
    \cline{1-3}
    112 & 980 & \(V_2,[V_1]\) \\
    \cline{1-3}
    \multicolumn{3}{l}{\multirow{2}{*}{\(9 \Big\downarrow [\mathbf{return}]\)}} & \\
    \multicolumn{3}{l}{} & \multirow{3}{*}{\(\underbrace{\dots \freebox \freebox \freebox \freebox \freebox}_\unsealed
      \! \underbrace{\stackrel{\stackrel{\SP}{\downarrow}}{\objbox} \!\! \objbox \objbox \objbox \objbox}_\object
      \! \underbrace{\pubbox \pubbox \pubbox \dots}_\public
      ~ \stackrel{\mathtt{a0}}{\pubbox} ~ \stackrel{\mathtt{a1}}{\freebox}
 %     ~ \stackrel{\mathtt{s0}}{\sealbox}
      \)}
    \\
    \cline{1-3}
    20 & 980  & \(V_1, \emplist\) &
    \\
    \cline{1-3}
    \multicolumn{2}{l}{} \\
  \end{tabular}
  \caption{Execution of example up through the return from {\tt f}. In right-hand diagram, addresses increase to the right and
  boxes represent 4-byte words.}
\label{fig:exec1}
\end{figure*}
%
At step 5, the current principal's record is pushed onto the inactive list.
Its return target is the return address of the call,
and the stack pointer target is the stack pointer at the moment of call.
The callee's view is updated from the caller's such that all \(\object\) locations
become \(\sealed\). (In more complex settings, objects that are intentionally passed
to the callee will not get sealed, but for now we assume no sharing of memory between activations.)
%For registers, \(\overline{\reg_{args}}\) tells us which registers are used as arguments, 
% in this case just {\tt a0}. \apt{We've lost the explanation of the arguments to call, and they are used inconsistent now too, so this is utterly mysterious.}
\apt{needs review}
{\tt f} does not take any arguments; if it did, any registers containing them would be 
mapped to \(\public\), while any non-argument, caller-saved
registers are mapped to \(\unsealed\). All callee-save registers remain \(\sealed\) for all calls.
At step 9, {\tt f} returns, and the topmost inactive view, that of {\tt main}, is restored.

We now show how this security semantics can be used define notions of confidentiality,
integrity, and correct control flow in such a way that many classes of
bad behavior, including the attacks in Figure~\ref{fig:f}, are
detected as security violations.

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential. In Figure \ref{subfig:WBCF}, the attacker adds
16 to the return address and then returns, thus bypassing the {\tt if}-test in the code and outputting
{\tt secret}.
In Figure \ref{subfig:WBCF2}, the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}
\end{center}

@@

\noindent
{\tt main}'s attempt to read {\tt sensitive} will instead
read part of the return address, and its attempt to output
{\tt res} will instead output {\tt secret}.

Before the call, the program counter is 16 and the stack pointer is 980.
So we define a predicate on states that should hold just after the return:
\(\ret\ \mach \triangleq \mach[\PCname] = 20 \wedge \mach[\SP] = 980\).
%
We can identify the point just after the return (if a return occurs)
as the first state in which the pending call stack is smaller than it was
just after the call.
WBCF requires that if \(\mach\) is the state at that point, then \(\ret ~ \mach\) holds.
%For nested calls, where the pending stack is initially larger, the same principle
%applies: \(\ret ~ \mach\) must hold the next time the pending stack is the same size or smaller.

% Even absent other
% kinds of data protection, the stack pointer {\it must} be restored
% for the program to behave predictably.

@@

Like WBCF, stack integrity defines a condition at the call that must hold upon
return. This time the condition applies to all of the memory in the function's 
frame. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labels it \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, any \(\sealed\) elements
are identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

@@

To characterize this situation in a property, we borrow the idea of \emph{variant} states
from the theory of non-interference. For a set of elements \(\components\),
a pair of states \(\mach\) and \(\nach\) are {\em \(\components\)-variants} if
their values an only disagree on elements in \(\components\).
We say that the elements of \(\components\) are \emph{irrelevant}
in \(\mach\) if they can be replaced by arbitrary other values without changing the
observable behavior of the machine. All other elements are \emph{relevant}.

(This story is slightly over-simplified. If an enforcement mechanism maintains
additional state associated with elements, such as tags, we don't want that
state to vary; only the payloads should. Formal definitions of variants and
relevance that incorporate this wrinkle are given in \cref{sec:props}.)

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{variants.png}
  \caption{Integrity Violation\apt{Needs a machine-drawn version, as do the rest. Use colors consistently. No tags!}}
  \label{fig:variant}
\end{figure}

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to WBCF, we define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43. As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

@@

\apt{We may need a new section that describes how a tag-based system would enforce the properties
  described so far.}

@@

The system we've modeled so far has been very simple, but our properties
support an important dimension of generality: adding features. We now
briefly introduce the features of our primary testing machine.
